---
title: "Working with larger-than-memory data"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{Working with larger-than-memory data}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(tolower(Sys.getenv("NOT_CRAN")), "true"),
  out.width = "100%"
)

```
# Larger-than-memory Data

Many of data sets of Brazilian censuses are too big to load in users' RAM memory. To avoid this problem, **censobr** works with files saved in `.parquet` format and, by default, returns an [Arrow table](https://arrow.apache.org/docs/r/articles/arrow.html#tabular-data-in-arrow).

There are three really simple alternative ways to work with Arrow tables in R without loading the full data to memory. We cover these three alternative approaches in this vignette.

First, let's read the 2010 2010 mortality data, which we'll use throughout vignette.

```{r warning = FALSE}
library(censobr)

# read 2010 mortality data
df <- censobr::read_mortality(year = 2010,
                              add_labels = 'pt')
```

## 1.  {`dplyr`} 

Because of the seamless integration between {`arrow`} and  {`dplyr`} , Arrow tables can be analyzed like a regular `data.frame` using the {`dplyr`} syntax. Note that we need to add a `collect()` call at the end of our query to retrieve data into a local tibble.

```{r warning = FALSE}
library(dplyr)

# Filter deaths of men in the state of Rio de Janeiro
rio <- df |>
      filter(V0704 == 'Masculino' & abbrev_state == 'RJ')

head(rio) |> collect()

```

## 2. {`duckdb`}

{`duckdb`} is another powerful library to work with larger-than-memory data in R through database interface. Here, one can use {`duckdb`} in three ways using a bit of `SQL` syntax.


### 2.1 Combining {`duckdb`} & {`dplyr`}

One easy option is to combine {`duckdb`} & {`dplyr`}. Note here that first you need to convert the Arrow table into a DuckDB table with `arrow::to_duckdb()`. Also note that the you need to use a bit of `SQL` syntax inside the dplyr call. Using the same example as above:

```{r warning = FALSE}
library(duckdb)
library(DBI)
library(arrow)

# Filter deaths of men in the state of Rio de Janeiro
rio1 <- df |>
        arrow::to_duckdb() |>
        filter(sql("V0704 LIKE '%Masculino%' AND abbrev_state = 'RJ'"))

head(rio1) |> collect()

```

### 2.2 Combining {`duckdb`} &  `SQL`

The other alternative is to combine {`duckdb`} with {`DBI`} using database interface and `SQL` syntax.


```{r warning = FALSE}
library(DBI)

# create a (virtual) database
db_tbl <- df |>
          to_duckdb(table_name = "mortality_2010_tbl")

# create databse connection
con <- db_tbl$src$con

# Filter deaths of men in the state of Rio de Janeiro
rio2 <- DBI::dbGetQuery(con, "SELECT * FROM 'mortality_2010_tbl' WHERE V0704 LIKE '%Masculino%' AND abbrev_state = 'RJ'")

head(rio2)

```
### 2.3 {`duckplyr`} 

A third alternative is the brand new {`duckplyr`} package. This is a really promising approach. However, this library is heavily under development and it still does not cover many of the {`dplyr`} verbs. See [here](https://duckdb.org/2024/04/02/duckplyr.html).

```{r warning = FALSE, eval = FALSE}
library(duckplyr)

rio3 <- df |>
        duckplyr::filter(V0704 == 'Masculino' & abbrev_state == 'RJ')

head(rio3) |> collect()
```
